Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2021-09-12 paper/latex/rmt.tex
wip

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous [1]–[4]. This complexity suggests fMRI is a
candidate to be studied using Random Matrix Theory (RMT), a set of mathematical tools originally
developed some 50 years ago to solve complex problems in nuclear physics [5], [6]. In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

\begin{sloppypar}
These universal properties have been observed in diverse phenomena. In small-scale systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems [5]–[11], and at larger scales, RMT has been applied to atmospheric
physics [12], stock cross-correlations [13], social networks [14], random networks [15],
network-formation in liquids [16], [17] and amorphous clusters [18]–[21]. Within biological systems,
RMT has also been used to successfully model aspects of  amino acid functional relationships [22],
synchrony in epileptic seizures [23], and in protein-protein interactions both in different species
[24] and breast cancer [25].
\end{sloppypar}

In recent years, RMT analyses have been applied to study brain functioning.  The earliest  study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) [26].  Somewhat more recently, RMT was used
to evaluate the quality of whole brain features extracted from fMRI data [27], [28]. RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising [29]–[31].

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states [32], between subjects with and without ADHD [33], and
between pain and non-pain states [34]. Across these three studies, the spectra of resting or
low-attention states exhibited properties closest to the GOE.

@@

However, since previous studies employing RMT for feature extraction [33], [34] have partially
interpreted their findings with respect to attentional processes, we have elected to use datasets
that involve attention in some way. This includes measures of attention or vigilance, or comparisons
between a condition involving more attention and focus (e.g. a task state) or conditions which
should involve considerably less attention or focus (e.g. resting-state).

@@

\subsubsection{Dataset 1}
Dataset 1 (referred to in Fig. and tables with the label "PSYCH") has detailed and extensive
information available in  [37]. Subjects completed a battery of psychological measurements, and
various resting-state fMRI (rs-fMRI) images were obtained. Only the whole-brain 7-Tesla rs-fMRI
scans were used for the current paper, to make the results comparable to the other included
datasets. For the purpose of keeping with the "attention" theme of the analyses in this paper, three
pairings of subgroups were constructed via z-score splits of the various sum-scores of scale items
involving concepts closely related to attention.

\paragraph{Trait Attention}
The PANAS-X [38] is a 9-point Likert-type self-report measure that queries the extent to which
subjects agree particular affective terms have applied to them over a recent time period (e.g. a few
weeks). An ad hoc (i.e. face-valid) sum-score was constructed from the scale items: "attentive",
"alert", "interested", "concentrating", and reverse-scored items "sluggish", "tired", "sleepy",
"drowsy". Subjects with a z-score greater / less than zero on this sum-score were placed in the high
/ low "trait attention" category

\paragraph{Task Attention}
The Conjunctive Continuous Performance Task (CCPT [39]) is a behavioural measure of sustained
attention where subjects must react only if they see a red square among other different coloured
figures. For this paper, a sum score was constructed by subtracting the number of false positive and
false negative responses from the number of success responses, and subjects with a z-score greater /
less than zero on this scale were placed in high / low "task attention" category.

\paragraph{Vigilance}
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. "my thoughts involved other people", "I thought about something positive" (translated
from German) [40]. One question assesses alertness ("I was fully awake" / "war ich vollkommen wach")
and was used to create high / low "vigilance" conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in Fig. and tables with the label "OSTEO") is detailed in [41]. Subjects
included healthy controls ("nopain" condition) and individuals with knee osteoarthritis ("pain"
condition) treated for 2 weeks with either placebo ("pain" condition) or duloxetine ("duloxetine"
condition). An additional grouping was also created with both the placebo and duloxetine data
("allpain" condition), allowing for six subgroup comparisons. All subjects were imaged with
whole-brain rs-fMRI. Scan parameters are documented in Table I. The use of rs-fMRI and RMT for
examining pain makes this dataset complementary to [34].

\subsubsection{Dataset 3}
Dataset 3  (referred to here with the label "REFLECT", due to the "reflective” nature of the
cognitive tasks) is described in full detail by DuPre, Luh and Spreng [42]. Both multi-echo rs- and
task-fMRI scans were conducted with scan parameters summarized in Table II. Tasks involved viewing
various affective images (see [43] for details) and performing tasks related to autobiographical
memory, prospection, or theory-of-mind reasoning. Continuing with the theme of broadly comparing
high versus low attention throughout this paper, the different tasks were simply treated generically
as "attending", in contrast to the non-attentive resting-state scans. This resulted in four task
state scans and one resting-state scan per subject.

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the "summed" condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast [44]. In the "interleave" condition, echoes were interleaved
at each voxel: that is, if we denote the voxel's ith signal value from echo j as , then the
interleaved voxel of the new image has the time-series
\([t_{1,1}, t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

@@

\subsubsection{Dataset 4}
Dataset 4 (referred to in Fig. and tables with the label "PARK" or "PARKINSONS") is documented and
described in detail in [45]. Scan parameters are documented in Table I. Subjects with non-demented
Parkinson’s Disease (PD) and healthy controls performed a number of repetitions of the Attention
Network Task (ANT) [46]. Subgroup details are available in Table I. As the ANT task demands high
attention, it is not as straightforward to create "high" or "low" attention groups. Instead, since
attention presumably is high and varies little in this dataset, and subgroups instead differ by
virtue of the presence or absence of PD, any results here serve as a useful comparison point for the
results from studies that should clearly involve attentional differences. This dataset also included
extensively preprocessed scans, which were included in analyses to examine the effects of
preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}
Dataset 5 (referred to in Fig. and tables with the label "LEARN") is documented and described in
detail in [47]. In this study, both rs-fMRI and task fMRI scans were available for all subjects,
allowing for within-subject comparisons. The complex task involved multiple phases, wherein subjects
were exposed to named, black-and-white outlined satellite images related by family-resemblance of
the components (see Fig. 1 of [47]). After this learning task, subjects were tested on their memory
of the various learned features. Scans corresponding to both the initial learning and subsequent
testing phase were included in the high-attention "task" subgroup for the purposes of the analysis
of this paper. Resting state scans were included in the low-attention "rest" subgroup.

\subsection{Preprocessing}
Some consideration should be given to whether it is appropriate to apply standard preprocessing
steps before extracting RMT features, as preprocessing steps should be developed, tested, and
employed always with respect to a set of inferential goals and a specific analytic framework.
However, these preprocessing steps are increasingly recognized to have complex effects on fMRI
analyses [48]. For example, spatial smoothing has clear benefits when employing a general linear
model [49]–[51], but is less motivated in functional network analyses with small ROIs, where the
effect of smoothing can be considerable [50], [52], [53], with these effects interacting complexly
with factors like field strength and image resolution [54]. Similarly, image registration is usually
required to allow comparisons across (or even within) subjects, but the choice of registration
strategy can have dramatic effects on functional network analyses [55].

>>>>

Whether in the presence of experimental stimuli, or the relative absence, as in a resting state,
complex functional connectivity networks are ubiquitous
(\cite{bucknerBrainDefaultNetwork2008,foxCoverHumanBrain2005,gonzalez-castilloTaskbasedDynamicFunctional2018,hermundstadStructuralFoundationsRestingstate2013}).
This complexity suggests fMRI is a candidate to be studied using Random Matrix Theory (RMT), a set
of mathematical tools originally developed some 50 years ago to solve complex problems in nuclear
physics (\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}). In various
physical systems where nuclei can have different possible states, precisely modeling the entirety of
the interactions between nuclei is often intractable. For such large, complex systems, the
eigenvalues (or spectra) efficiently summarize the totality of the interactions between components
of the system, and RMT describes the expected behaviour of such eigenvalues. By analysing the
spectrum statistically, complex systems can be investigated by comparison to the universal
properties predicted by RMT.

\begin{sloppypar}
These universal properties have been observed in diverse phenomena. In small-scale systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems \cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has been applied to atmospheric
physics (\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations \cite{plerouRandomMatrixApproach2002} , social networks \cite{jalanUncoveringRandomnessSuccess2014}, random networks \cite{bandyopadhyayUniversalityComplexNetworks2007},
network-formation in liquids (\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009})  and amorphous clusters (\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}). Within biological systems,
RMT has also been used to successfully model aspects of  amino acid functional relationships (\cite{bhadolaTargetingFunctionalMotifs2016}),
synchrony in epileptic seizures (\cite{osorioPhasesynchronizationRandommatrixBased2011}), and in protein-protein interactions both in different species (\cite{agrawalQuantifyingRandomnessProtein2014}) and breast cancer (\cite{raiRandomnessPreservedPatterns2015}).
\end{sloppypar}

In recent years, RMT analyses have been applied to study brain functioning.  The earliest  study
demonstrated that spectra of the correlations between electroencephalographic (EEG) signals closely
resemble those of the Gaussian Orthogonal Ensemble (GOE) (\cite{sebaRandomMatrixAnalysis2003}).
Somewhat more recently, RMT was used to evaluate the quality of whole brain features extracted from
fMRI data (\cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}). RMT has also
been used in diffusion MRI to aid in the selection of the number of components to employ in
principal-component reduction analysis and denoising
(\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}).

Finally, RMT has been used in ROI-based fMRI functional-connectivity studies to investigate
differences between rest and task states (\cite{wangSpectralPropertiesTemporal2015}), between
subjects with and without ADHD (\cite{wangRandomMatrixTheory2016}), and between pain and non-pain
states (\cite{matharooSpontaneousBackpainAlters2020}). Across these three studies, the spectra of
resting or low-attention states exhibited properties closest to the GOE.

@@

However, since previous studies employing RMT for feature extraction
(\cite{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}) have partially interpreted
their findings with respect to attentional processes, we have elected to use datasets that involve
attention in some way. This includes measures of attention or vigilance, or comparisons between a
condition involving more attention and focus (e.g. a task state) or conditions which should involve
considerably less attention or focus (e.g. resting-state).

@@

\subsubsection{Dataset 1}
Dataset 1 (referred to in Fig. and tables with the label "PSYCH") has detailed and extensive
information available in \cite{gorgolewskiHighResolution7Tesla2015}. Subjects completed a battery of
psychological measurements, and various resting-state fMRI (rs-fMRI) images were obtained. Only the
whole-brain 7-Tesla rs-fMRI scans were used for the current paper, to make the results comparable to
the other included datasets. For the purpose of keeping with the "attention" theme of the analyses
in this paper, three pairings of subgroups were constructed via z-score splits of the various
sum-scores of scale items involving concepts closely related to attention.

\paragraph{Trait Attention}
The PANAS-X (\cite{watsonPANASXManualPositive1994}) is a 9-point Likert-type self-report measure
that queries the extent to which subjects agree particular affective terms have applied to them over
a recent time period (e.g. a few weeks). An ad hoc (i.e. face-valid) sum-score was constructed from
the scale items: "attentive", "alert", "interested", "concentrating", and reverse-scored items
"sluggish", "tired", "sleepy", "drowsy". Subjects with a z-score greater / less than zero on this
sum-score were placed in the high / low "trait attention" category

\paragraph{Task Attention}
The Conjunctive Continuous Performance Task (CCPT;
\cite{shalevConjunctiveContinuousPerformance2011}) is a behavioural measure of sustained attention
where subjects must react only if they see a red square among other different coloured figures. For
this paper, a sum score was constructed by subtracting the number of false positive and false
negative responses from the number of success responses, and subjects with a z-score greater / less
than zero on this scale were placed in high / low "task attention" category.

\paragraph{Vigilance}
The Mini New York cognition questionnaire assesses various elements of the subject's cognitive
state, e.g. "my thoughts involved other people", "I thought about something positive" (translated
from German) \cite{gorgolewskiCorrespondenceIndividualDifferences2014}. One question assesses
alertness ("I was fully awake" / "war ich vollkommen wach") and was used to create high / low
"vigilance" conditions in the same manner as the previous two splits.

\subsubsection{Dataset 2}
Dataset 2 (referred to in Fig. and tables with the label "OSTEO") is detailed in
\cite{tetreaultBrainConnectivityPredicts2016}. Subjects included healthy controls ("nopain"
condition) and individuals with knee osteoarthritis ("pain" condition) treated for 2 weeks with
either placebo ("pain" condition) or duloxetine ("duloxetine" condition). An additional grouping was
also created with both the placebo and duloxetine data ("allpain" condition), allowing for six
subgroup comparisons. All subjects were imaged with whole-brain rs-fMRI. Scan parameters are
documented in Table I. The use of rs-fMRI and RMT for examining pain makes this dataset
complementary to \cite{matharooSpontaneousBackpainAlters2020}.

\subsubsection{Dataset 3}
Dataset 3  (referred to here with the label "REFLECT", due to the "reflective” nature of the
cognitive tasks) is described in full detail by \cite{dupreMultiechoFMRIReplication2016}. Both
multi-echo rs- and task-fMRI scans were conducted with scan parameters summarized in Table II. Tasks
involved viewing various affective images (see \cite{sprengPatternsBrainActivity2010} for details)
and performing tasks related to autobiographical memory, prospection, or theory-of-mind reasoning.
Continuing with the theme of broadly comparing high versus low attention throughout this paper, the
different tasks were simply treated generically as "attending", in contrast to the non-attentive
resting-state scans. This resulted in four task state scans and one resting-state scan per subject.

In order to combine the multi-echo data into a single, usable matrix, two methods were employed. In
the "summed" condition, echoes were combined via summation, a method which has value in providing an
image with overall increased contrast (\cite{kunduMultiechoFMRIReview2017}). In the "interleave"
condition, echoes were interleaved at each voxel: that is, if we denote the voxel's ith signal value
from echo j as , then the interleaved voxel of the new image has the time-series \([t_{1,1},
t_{2,1}, t_{3,1}, \dots, t_{1,i}, t_{2,i}, t_{3,i}, \dots, t_{1,n}, t_{2,n}, t_{3,n}]\).

@@

\subsubsection{Dataset 4}
Dataset 4 (referred to in Fig. and tables with the label "PARK" or "PARKINSONS") is documented and
described in detail in \cite{madhyasthaDynamicConnectivityRest2015}. Scan parameters are documented
in Table I. Subjects with non-demented Parkinson’s Disease (PD) and healthy controls performed a
number of repetitions of the Attention Network Task (ANT)
(\cite{fanActivationAttentionalNetworks2005}). Subgroup details are available in Table I. As the ANT
task demands high attention, it is not as straightforward to create "high" or "low" attention
groups. Instead, since attention presumably is high and varies little in this dataset, and subgroups
instead differ by virtue of the presence or absence of PD, any results here serve as a useful
comparison point for the results from studies that should clearly involve attentional differences.
This dataset also included extensively preprocessed scans, which were included in analyses to
examine the effects of preprocessing on the extracted RMT features.

\subsubsection{Dataset 5}
Dataset 5 (referred to in Fig. and tables with the label "LEARN") is documented and described in
detail in \cite{schapiroHumanHippocampalReplay2018}. In this study, both rs-fMRI and task fMRI scans
were available for all subjects, allowing for within-subject comparisons. The complex task involved
multiple phases, wherein subjects were exposed to named, black-and-white outlined satellite images
related by family-resemblance of the components (see Fig. 1 of
\cite{schapiroHumanHippocampalReplay2018}). After this learning task, subjects were tested on their
memory of the various learned features. Scans corresponding to both the initial learning and
subsequent testing phase were included in the high-attention "task" subgroup for the purposes of the
analysis of this paper. Resting state scans were included in the low-attention "rest" subgroup.

\subsection{Preprocessing}
Some consideration should be given to whether it is appropriate to apply standard preprocessing
steps before extracting RMT features, as preprocessing steps should be developed, tested, and
employed always with respect to a set of inferential goals and a specific analytic framework.
However, these preprocessing steps are increasingly recognized to have complex effects on fMRI
analyses (\cite{parkerBenefitSliceTiming2019}). For example, spatial smoothing has clear benefits
when employing a general linear model
(\cite{miklEffectsSpatialSmoothing2008,alakorkkoEffectsSpatialSmoothing2017,worsleyAnalysisFMRITimeSeries1995}),
but is less motivated in functional network analyses with small ROIs, where the effect of smoothing
can be considerable
(\cite{alakorkkoEffectsSpatialSmoothing2017,goelmanMaximizingNegativeCorrelations2014,trianaEffectsSpatialSmoothing2020},
with these effects interacting complexly with factors like field strength and image resolution
(\cite{triantafyllouEffectSpatialSmoothing2006}). Similarly, image registration is usually required
to allow comparisons across (or even within) subjects, but the choice of registration strategy can
have dramatic effects on functional network analyses (\cite{magalhaesImpactNormalizationSegmentation2014}).
